---
title: "Lab 4: Transcriptomics"
author: "Catherine"
format: html
editor: visual
---

## <https://quarto.org>.

## 9.1.1 Question 1

##### Accessing the RNAseq data and sample data

```{r}
# Read the meta table 
meta_data <- read.table(
  system.file("extdata/param/targetsPE.txt", package = "systemPipeRdata"), 
  header = TRUE, 
  stringsAsFactors = FALSE
)

head(meta_data)

# Replace the filename
# Get the actual path to the FASTQ files directory
fq_path <- systemPipeRdata::pathList()$fastqdir

# Replace './data/' with the actual directory path in FileName1 and FileName2
meta_data$FileName1 <- sub("^\\./data/", paste0(fq_path, "/"), meta_data$FileName1)
meta_data$FileName2 <- sub("^\\./data/", paste0(fq_path, "/"), meta_data$FileName2)

# Verify the updated paths
head(meta_data[, c("FileName1", "FileName2")])
# Display the metadata table using DT package
library(DT)

# Render interactive table
datatable(
  meta_data, 
  options = list(scrollX = TRUE, pageLength = 5), 
  caption = "RNA-seq Sample Metadata Table"
)

```

## 9.2.1 Question 2

##### Read processing

```{r}
# Generate the QC report
library(Rqc)
library(ggplot2)

# Combine all forward and reverse file paths from your metadata table
all_files <- c(meta_data$FileName1, meta_data$FileName2)

# Run the overall Rqc quality assessment and generate the HTML report
qc_report <- rqc(path = dirname(all_files[1]), pattern = "fastq\\.gz$")
```

##### Generate 3 per-cycle Q-score box plot for files 1-12, 13-24, 25-36

```{r}
# Split the files into three equal groups of 12
files_1_12  <- all_files[1:12]
files_13_24 <- all_files[13:24]
files_25_36 <- all_files[25:36]

# Run rqcQA for each batch
qa_1 <- rqcQA(files_1_12)
qa_2 <- rqcQA(files_13_24)
qa_3 <- rqcQA(files_25_36)

# Generate per-cycle Q-score box plots with black and white theme
rqcCycleQualityBoxPlot(qa_1) + theme_bw()
rqcCycleQualityBoxPlot(qa_2) + theme_bw()
rqcCycleQualityBoxPlot(qa_3) + theme_bw()
```

##### Are there any samples with much longer reads than others?

- No, they share a uniform read length

##### What’s the shortest number of cycles?

- The shortest number of cycle is about 75 cycles

##### Generate 3 base call frequency plots for files 1-12, 13-24, 25-36

```{r}
# Generate base call frequency plots with black and white theme
rqcCycleBaseCallsPlot(qa_1) + theme_bw()
rqcCycleBaseCallsPlot(qa_2) + theme_bw()
rqcCycleBaseCallsPlot(qa_3) + theme_bw()
```

##### Does any sample have significant N calls?

- No, there is no signicicant N calls

## 9.2.2 Question 3

```{r}
dir.create("outputs/processed_fastq", recursive = TRUE, showWarnings = FALSE)

meta_data <- read.table(
  system.file("extdata/param/targetsPE.txt", package = "systemPipeRdata"),
  header = TRUE, stringsAsFactors = FALSE
)
fq_path <- systemPipeRdata::pathList()$fastqdir
meta_data$FileName1 <- sub("^\\./data/", paste0(fq_path, "/"), meta_data$FileName1)
meta_data$FileName2 <- sub("^\\./data/", paste0(fq_path, "/"), meta_data$FileName2)

meta_data$Proc1 <- file.path("outputs/processed_fastq",
  sub("\\.fastq\\.gz$", "_processed.fastq.gz", basename(meta_data$FileName1)))
meta_data$Proc2 <- file.path("outputs/processed_fastq",
  sub("\\.fastq\\.gz$", "_processed.fastq.gz", basename(meta_data$FileName2)))

for (i in seq_len(nrow(meta_data))) {
  unlink(c(meta_data$Proc1[i], meta_data$Proc2[i]))
  QuasR::preprocessReads(
    c(meta_data$FileName1[i], meta_data$FileName2[i]),
    c(meta_data$Proc1[i], meta_data$Proc2[i]),
    nBases = 1,
    truncateEndBases = 3,
    Lpattern = "GCCCGGGTAA",
    minLength = 40
  )
}
```

```{r}
library(ShortRead)

dir.create("outputs/paired_fastq", recursive = TRUE, showWarnings = FALSE)

meta_data$Pair1 <- file.path("outputs/paired_fastq", basename(meta_data$Proc1))
meta_data$Pair2 <- file.path("outputs/paired_fastq", basename(meta_data$Proc2))

for (i in seq_len(nrow(meta_data))) {
  r1 <- readFastq(meta_data$Proc1[i])
  r2 <- readFastq(meta_data$Proc2[i])

  # read IDs without the /1, /2 suffix or any text after a space
  id1 <- sub("[ /].*$", "", as.character(id(r1)))
  id2 <- sub("[ /].*$", "", as.character(id(r2)))

  common <- intersect(id1, id2)
  r1 <- r1[match(common, id1)]
  r2 <- r2[match(common, id2)]

  unlink(c(meta_data$Pair1[i], meta_data$Pair2[i]))
  writeFastq(r1, meta_data$Pair1[i], compress = TRUE)
  writeFastq(r2, meta_data$Pair2[i], compress = TRUE)
}


```

## 9.3.3 Question 4: Aligning the reads

```{r}
#| warning: false
library(stringr)

# Create output directories
dir.create(
  "outputs/sam_files",
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  "outputs/bam_files",
  recursive = TRUE,
  showWarnings = FALSE
)

# Loop through the metadata table
for (i in seq_along(meta_data$FileName1)) {
  
  sample_name <- meta_data$SampleName[i]
  
  # Processed paired-end FASTQ files
  fastq_file1 <- meta_data$Pair1[i]
  fastq_file2 <- meta_data$Pair2[i]
  
  # Output files
  sam_output <- file.path(
    "outputs/sam_files",
    paste0(sample_name, ".sam")
  )
  
  bam_output <- file.path(
    "outputs/bam_files",
    paste0(sample_name, ".bam")
  )
  
  sorted_bam_output <- file.path(
    "outputs/bam_files",
    paste0(sample_name, "_sorted.bam")
  )
  
  log_output <- file.path(
    "outputs/bam_files",
    paste0(sample_name, "_hisat2.log")
  )
  
  message("Aligning ", sample_name, "...")
  
  # Run HISAT2
  hisat2_result <- system2(
    command = "hisat2",
    args = c(
      "-x", "outputs/hisat2_index/tair10_1_index",
      "-1", fastq_file1,
      "-2", fastq_file2,
      "-p", "8",
      "-S", sam_output
    ),
    stdout = TRUE,
    stderr = TRUE
  )
  
  # Save HISAT2 statistics
  writeLines(
    hisat2_result,
    con = log_output
  )
  
  # Convert SAM to BAM
  system2(
    "samtools",
    args = c(
      "view",
      "-bS",
      sam_output,
      "-o",
      bam_output
    )
  )
  
  # Sort BAM
  system2(
    "samtools",
    args = c(
      "sort",
      bam_output,
      "-o",
      sorted_bam_output
    )
  )
  
  # Index sorted BAM
  system2(
    "samtools",
    args = c(
      "index",
      sorted_bam_output
    )
  )
  
  message("Finished ", sample_name)
}

```

## 9.3.4.1 Question 5: alignment stats

```{r}
library(DT)
library(dplyr)
library(ggplot2)
library(stringr)

# Directory containing HISAT2 log files
hisat2_logs_dir <- "outputs/bam_files"

# Find all HISAT2 log files
log_files <- list.files(
  hisat2_logs_dir,
  pattern = "_hisat2\\.log$",
  full.names = TRUE
)

# Check that all expected log files are present
if (length(log_files) != 18) {
  stop(
    paste(
      "Expected 18 HISAT2 log files, but found",
      length(log_files)
    )
  )
}

# Extract the overall alignment percentage from each log
align_list <- lapply(log_files, function(log_file) {
  
  # Read the log file
  log_text <- readLines(log_file, warn = FALSE)
  
  # Find the line containing the overall alignment rate
  alignment_line <- log_text[
    grepl(
      "overall alignment rate",
      log_text,
      ignore.case = TRUE
    )
  ]
  
  # Stop with a useful message if the alignment rate is missing
  if (length(alignment_line) == 0) {
    stop(
      paste(
        "Could not find overall alignment rate in:",
        basename(log_file)
      )
    )
  }
  
  # Extract the percentage from the alignment line
  percent <- as.numeric(
    str_extract(
      alignment_line[1],
      "[0-9]+\\.?[0-9]*(?=%)"
    )
  )
  
  # Extract sample name from log filename
  sample_name <- sub(
    "_hisat2\\.log$",
    "",
    basename(log_file)
  )
  
  data.frame(
    samplename = sample_name,
    percent_aligned = percent
  )
})

# Combine results into one data frame
align_df <- bind_rows(align_list)

# Sort samples according to the original metadata order
align_df$samplename <- factor(
  align_df$samplename,
  levels = meta_data$SampleName
)

align_df <- align_df %>%
  arrange(samplename)

# Convert sample name back to character
align_df$samplename <- as.character(align_df$samplename)

# Display alignment results
print(align_df)

# Interactive table
datatable(
  align_df,
  options = list(
    scrollX = TRUE,
    pageLength = 6
  ),
  caption = "HISAT2 Read Alignment Percentages per Sample"
)

# Boxplot of alignment percentages
ggplot(
  align_df,
  aes(y = percent_aligned)
) +
  geom_boxplot(
    fill = "steelblue",
    alpha = 0.7,
    outlier.color = "red",
    outlier.shape = 16
  ) +
  theme_bw() +
  labs(
    title = "Distribution of Read Alignment Percentages",
    y = "Percent Aligned (%)",
    x = ""
  ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )

```

## 9.4.1 Question 6: the count table

```{r}
library(DT)
library(dplyr)
library(Rsubread)

bfiles <- list.files("outputs/bam_files", pattern = "_sorted.bam$", full.names = TRUE)

gene_count_list <- Rsubread::featureCounts(
  files = bfiles,
  annot.ext = "data/GCF_000001735.4_TAIR10.1_genomic.gtf",
  isGTFAnnotationFile = TRUE,
  allowMultiOverlap = FALSE,
  isPairedEnd = TRUE, nthreads = 8,
  minMQS = 10,
  GTF.featureType = "exon",
  GTF.attrType = "gene_id"
)

# Assign the count matrix to count_table
count_table <- gene_count_list$counts

# Remove "_sorted.bam" from column names
colnames(count_table) <- sub(
  "_sorted\\.bam$",
  "",
  colnames(count_table)
)

# Keep only genes with at least one read mapped
# across all samples
count_table <- count_table[
  rowSums(count_table) > 0,
]

# Display the count table as an interactive table
datatable(
  count_table,
  options = list(
    scrollX = TRUE,
    pageLength = 10
  ),
  caption = "Gene-Level Read Count Table"
)
```