---
title: "Lab 4: Transcriptomics"
author: "Catherine"
format: html
editor: visual
---

## <https://quarto.org>.

## 9.1.1 Question 1

##### Accessing the RNAseq data and sample data

```{r}
# Read the meta table 
meta_data <- read.table(
  system.file("extdata/param/targetsPE.txt", package = "systemPipeRdata"), 
  header = TRUE, 
  stringsAsFactors = FALSE
)

head(meta_data)

# Replace the filename
# Get the actual path to the FASTQ files directory
fq_path <- systemPipeRdata::pathList()$fastqdir

# Replace './data/' with the actual directory path in FileName1 and FileName2
meta_data$FileName1 <- sub("^\\./data/", paste0(fq_path, "/"), meta_data$FileName1)
meta_data$FileName2 <- sub("^\\./data/", paste0(fq_path, "/"), meta_data$FileName2)

# Verify the updated paths
head(meta_data[, c("FileName1", "FileName2")])
# Display the metadata table using DT package
library(DT)

# Render interactive table
datatable(
  meta_data, 
  options = list(scrollX = TRUE, pageLength = 5), 
  caption = "RNA-seq Sample Metadata Table"
)

```

## 9.2.1 Question 2

##### Read processing

```{r}
# Generate the QC report
library(Rqc)
library(ggplot2)

# Combine all forward and reverse file paths from your metadata table
all_files <- c(meta_data$FileName1, meta_data$FileName2)

# Run the overall Rqc quality assessment and generate the HTML report
qc_report <- rqc(path = dirname(all_files[1]), pattern = "fastq\\.gz$")
```

##### Generate 3 per-cycle Q-score box plot for files 1-12, 13-24, 25-36

```{r}
# Split the files into three equal groups of 12
files_1_12  <- all_files[1:12]
files_13_24 <- all_files[13:24]
files_25_36 <- all_files[25:36]

# Run rqcQA for each batch
qa_1 <- rqcQA(files_1_12)
qa_2 <- rqcQA(files_13_24)
qa_3 <- rqcQA(files_25_36)

# Generate per-cycle Q-score box plots with black and white theme
rqcCycleQualityBoxPlot(qa_1) + theme_bw()
rqcCycleQualityBoxPlot(qa_2) + theme_bw()
rqcCycleQualityBoxPlot(qa_3) + theme_bw()
```

##### Are there any samples with much longer reads than others?

- No, they share a uniform read length

##### What’s the shortest number of cycles?

- The shortest number of cycle is about 75 cycles

##### Generate 3 base call frequency plots for files 1-12, 13-24, 25-36

```{r}
# Generate base call frequency plots with black and white theme
rqcCycleBaseCallsPlot(qa_1) + theme_bw()
rqcCycleBaseCallsPlot(qa_2) + theme_bw()
rqcCycleBaseCallsPlot(qa_3) + theme_bw()
```

##### Does any sample have significant N calls?

- No, there is no signicicant N calls

## 9.2.2 Question 3

```{r}
dir.create("outputs/processed_fastq", recursive = TRUE, showWarnings = FALSE)

meta_data <- read.table(
  system.file("extdata/param/targetsPE.txt", package = "systemPipeRdata"),
  header = TRUE, stringsAsFactors = FALSE
)
fq_path <- systemPipeRdata::pathList()$fastqdir
meta_data$FileName1 <- sub("^\\./data/", paste0(fq_path, "/"), meta_data$FileName1)
meta_data$FileName2 <- sub("^\\./data/", paste0(fq_path, "/"), meta_data$FileName2)

meta_data$Proc1 <- file.path("outputs/processed_fastq",
  sub("\\.fastq\\.gz$", "_processed.fastq.gz", basename(meta_data$FileName1)))
meta_data$Proc2 <- file.path("outputs/processed_fastq",
  sub("\\.fastq\\.gz$", "_processed.fastq.gz", basename(meta_data$FileName2)))

for (i in seq_len(nrow(meta_data))) {
  unlink(c(meta_data$Proc1[i], meta_data$Proc2[i]))
  QuasR::preprocessReads(
    c(meta_data$FileName1[i], meta_data$FileName2[i]),
    c(meta_data$Proc1[i], meta_data$Proc2[i]),
    nBases = 1,
    truncateEndBases = 3,
    Lpattern = "GCCCGGGTAA",
    minLength = 40
  )
}
```

```{r}
library(ShortRead)

dir.create("outputs/paired_fastq", recursive = TRUE, showWarnings = FALSE)

meta_data$Pair1 <- file.path("outputs/paired_fastq", basename(meta_data$Proc1))
meta_data$Pair2 <- file.path("outputs/paired_fastq", basename(meta_data$Proc2))

for (i in seq_len(nrow(meta_data))) {
  r1 <- readFastq(meta_data$Proc1[i])
  r2 <- readFastq(meta_data$Proc2[i])

  # read IDs without the /1, /2 suffix or any text after a space
  id1 <- sub("[ /].*$", "", as.character(id(r1)))
  id2 <- sub("[ /].*$", "", as.character(id(r2)))

  common <- intersect(id1, id2)
  r1 <- r1[match(common, id1)]
  r2 <- r2[match(common, id2)]

  unlink(c(meta_data$Pair1[i], meta_data$Pair2[i]))
  writeFastq(r1, meta_data$Pair1[i], compress = TRUE)
  writeFastq(r2, meta_data$Pair2[i], compress = TRUE)
}


```

## 9.3.3 Question 4: Aligning the reads

```{r}
#| warning: false
library(stringr)

# Create output directories
dir.create(
  "outputs/sam_files",
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  "outputs/bam_files",
  recursive = TRUE,
  showWarnings = FALSE
)

# Loop through the metadata table
for (i in seq_along(meta_data$FileName1)) {
  
  sample_name <- meta_data$SampleName[i]
  
  # Processed paired-end FASTQ files
  fastq_file1 <- meta_data$Pair1[i]
  fastq_file2 <- meta_data$Pair2[i]
  
  # Output files
  sam_output <- file.path(
    "outputs/sam_files",
    paste0(sample_name, ".sam")
  )
  
  bam_output <- file.path(
    "outputs/bam_files",
    paste0(sample_name, ".bam")
  )
  
  sorted_bam_output <- file.path(
    "outputs/bam_files",
    paste0(sample_name, "_sorted.bam")
  )
  
  log_output <- file.path(
    "outputs/bam_files",
    paste0(sample_name, "_hisat2.log")
  )
  
  message("Aligning ", sample_name, "...")
  
  # Run HISAT2
  hisat2_result <- system2(
    command = "hisat2",
    args = c(
      "-x", "outputs/hisat2_index/tair10_1_index",
      "-1", fastq_file1,
      "-2", fastq_file2,
      "-p", "8",
      "-S", sam_output
    ),
    stdout = TRUE,
    stderr = TRUE
  )
  
  # Save HISAT2 statistics
  writeLines(
    hisat2_result,
    con = log_output
  )
  
  # Convert SAM to BAM
  system2(
    "samtools",
    args = c(
      "view",
      "-bS",
      sam_output,
      "-o",
      bam_output
    )
  )
  
  # Sort BAM
  system2(
    "samtools",
    args = c(
      "sort",
      bam_output,
      "-o",
      sorted_bam_output
    )
  )
  
  # Index sorted BAM
  system2(
    "samtools",
    args = c(
      "index",
      sorted_bam_output
    )
  )
  
  message("Finished ", sample_name)
}

```

## 9.3.4.1 Question 5: alignment stats

```{r}
library(DT)
library(dplyr)
library(ggplot2)
library(stringr)

# Directory containing HISAT2 log files
hisat2_logs_dir <- "outputs/bam_files"

# Find all HISAT2 log files
log_files <- list.files(
  hisat2_logs_dir,
  pattern = "_hisat2\\.log$",
  full.names = TRUE
)

# Check that all expected log files are present
if (length(log_files) != 18) {
  stop(
    paste(
      "Expected 18 HISAT2 log files, but found",
      length(log_files)
    )
  )
}

# Extract the overall alignment percentage from each log
align_list <- lapply(log_files, function(log_file) {
  
  # Read the log file
  log_text <- readLines(log_file, warn = FALSE)
  
  # Find the line containing the overall alignment rate
  alignment_line <- log_text[
    grepl(
      "overall alignment rate",
      log_text,
      ignore.case = TRUE
    )
  ]
  
  # Stop with a useful message if the alignment rate is missing
  if (length(alignment_line) == 0) {
    stop(
      paste(
        "Could not find overall alignment rate in:",
        basename(log_file)
      )
    )
  }
  
  # Extract the percentage from the alignment line
  percent <- as.numeric(
    str_extract(
      alignment_line[1],
      "[0-9]+\\.?[0-9]*(?=%)"
    )
  )
  
  # Extract sample name from log filename
  sample_name <- sub(
    "_hisat2\\.log$",
    "",
    basename(log_file)
  )
  
  data.frame(
    samplename = sample_name,
    percent_aligned = percent
  )
})

# Combine results into one data frame
align_df <- bind_rows(align_list)

# Sort samples according to the original metadata order
align_df$samplename <- factor(
  align_df$samplename,
  levels = meta_data$SampleName
)

align_df <- align_df %>%
  arrange(samplename)

# Convert sample name back to character
align_df$samplename <- as.character(align_df$samplename)

# Display alignment results
print(align_df)

# Interactive table
datatable(
  align_df,
  options = list(
    scrollX = TRUE,
    pageLength = 6
  ),
  caption = "HISAT2 Read Alignment Percentages per Sample"
)

# Boxplot of alignment percentages
ggplot(
  align_df,
  aes(y = percent_aligned)
) +
  geom_boxplot(
    fill = "steelblue",
    alpha = 0.7,
    outlier.color = "red",
    outlier.shape = 16
  ) +
  theme_bw() +
  labs(
    title = "Distribution of Read Alignment Percentages",
    y = "Percent Aligned (%)",
    x = ""
  ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )

```

## 9.4.1 Question 6: the count table

```{r}
library(DT)
library(dplyr)
library(Rsubread)

bfiles <- list.files("outputs/bam_files", pattern = "_sorted.bam$", full.names = TRUE)

gene_count_list <- Rsubread::featureCounts(
  files = bfiles,
  annot.ext = "data/GCF_000001735.4_TAIR10.1_genomic.gtf",
  isGTFAnnotationFile = TRUE,
  allowMultiOverlap = FALSE,
  isPairedEnd = TRUE, nthreads = 8,
  minMQS = 10,
  GTF.featureType = "exon",
  GTF.attrType = "gene_id"
)

# Assign the count matrix to count_table
count_table <- gene_count_list$counts

# Remove "_sorted.bam" from column names
colnames(count_table) <- sub(
  "_sorted\\.bam$",
  "",
  colnames(count_table)
)

# Keep only genes with at least one read mapped
# across all samples
count_table <- count_table[
  rowSums(count_table) > 0,
]

# Display the count table as an interactive table
datatable(
  count_table,
  options = list(
    scrollX = TRUE,
    pageLength = 10
  ),
  caption = "Gene-Level Read Count Table"
)
```



## 9.5.3.2 Question 7

```{r}
library(DESeq2)
library(stringr)
library(tidyr)
library(ggplot2)
library(dplyr)
library(ape)

# Create DESeq2 dataset
dds2 <- DESeqDataSetFromMatrix(
  countData = count_table,
  colData = meta_data,
  design = ~ Factor
)

# Run DESeq2
dds2 <- DESeq(dds2)

# Store DESeq2 dataset for Question 7
dds2_results <- dds2

# Read the comparisons from the target file
comp <- systemPipeR::readComp(
  system.file(
    "extdata/param/targetsPE.txt",
    package = "systemPipeRdata"
  )
)

comp[[1]]

# Split comparisons
pairs <- strsplit(comp[[1]], "-")

# Function to count significantly DE genes
filter_and_count <- function(
  res_obj,
  comparison_name,
  fc_threshold = 2,
  padj_threshold = 0.2
) {
  
  # Remove genes with missing values
  res_filtered <- res_obj[
    !is.na(res_obj$padj) &
    !is.na(res_obj$log2FoldChange),
  ]
  
  # Keep genes with:
  # 1. adjusted p-value < 0.2
  # 2. absolute fold change >= 2
  sig_genes <- res_filtered[
    res_filtered$padj < padj_threshold &
    abs(res_filtered$log2FoldChange) >= log2(fc_threshold),
  ]
  
  data.frame(
    Comparison = comparison_name,
    Up_regulated = sum(sig_genes$log2FoldChange > 0),
    Down_regulated = sum(sig_genes$log2FoldChange < 0)
  )
}

# Calculate results for each comparison
results_summary2 <- do.call(
  rbind,
  lapply(pairs, function(p) {
    
    res <- DESeq2::results(
      dds2_results,
      contrast = c("Factor", p[1], p[2]),
      alpha = 0.2
    )
    
    filter_and_count(
      res,
      paste(p[1], "vs", p[2])
    )
  })
)

# Display summary
print(results_summary2)

# Reshape data for plotting
plot_data2 <- results_summary2 %>%
  pivot_longer(
    cols = c(Up_regulated, Down_regulated),
    names_to = "Regulation",
    values_to = "Count"
  ) %>%
  mutate(
    Regulation = factor(
      Regulation,
      levels = c("Up_regulated", "Down_regulated")
    )
  )

# Plot
ggplot(
  plot_data2,
  aes(
    x = Comparison,
    y = Count,
    fill = Regulation
  )
) +
  geom_bar(
    stat = "identity",
    position = "stack"
  ) +
  coord_flip() +
  labs(
    title = "Differentially Expressed Genes by Comparison",
    subtitle = "Adjusted p-value < 0.2 and fold change >= 2",
    x = "Comparison",
    y = "Number of Genes",
    fill = "Regulation"
  ) +
  theme_minimal()
```

## 9.45.3.2 Question 8
```{r}
#| label: q8-heatmap
#| fig-cap: "Clustered heatmap of significant differentially expressed genes."
#| fig-width: 10
#| fig-height: 8

library(DESeq2)
library(pheatmap)

# Get normalized counts
norm_counts <- counts(
  dds2_results,
  normalized = TRUE
)

# Get differential expression results for the three comparisons
res_vir_mock <- DESeq2::results(
  dds2_results,
  contrast = c("Factor", "V1", "M1"),
  alpha = 0.05
)

res_avr_mock <- DESeq2::results(
  dds2_results,
  contrast = c("Factor", "A1", "M1"),
  alpha = 0.05
)

res_vir_avr <- DESeq2::results(
  dds2_results,
  contrast = c("Factor", "V1", "A1"),
  alpha = 0.05
)

# Identify significant genes from each comparison
sig_vir_mock <- rownames(res_vir_mock)[
  !is.na(res_vir_mock$padj) &
    res_vir_mock$padj < 0.05
]

sig_avr_mock <- rownames(res_avr_mock)[
  !is.na(res_avr_mock$padj) &
    res_avr_mock$padj < 0.05
]

sig_vir_avr <- rownames(res_vir_avr)[
  !is.na(res_vir_avr$padj) &
    res_vir_avr$padj < 0.05
]

# Combine significant genes from all three comparisons
sig_genes <- unique(
  c(
    sig_vir_mock,
    sig_avr_mock,
    sig_vir_avr
  )
)

# Keep only genes present in the normalized count matrix
sig_genes <- intersect(
  sig_genes,
  rownames(norm_counts)
)

# Extract normalized counts for significant DEGs
heatmap_data <- norm_counts[
  sig_genes,
  ,
  drop = FALSE
]

# Log2 transform the normalized counts
heatmap_data_log <- log2(
  heatmap_data + 1
)

# Row-wise Z-score normalization
heatmap_z <- t(
  scale(
    t(heatmap_data_log)
  )
)

# Remove genes with NA values after scaling
heatmap_z <- heatmap_z[
  complete.cases(heatmap_z),
  ,
  drop = FALSE
]

# Create sample annotation
sample_annotation <- data.frame(
  Factor = as.character(dds2_results$Factor)
)

rownames(sample_annotation) <- colnames(heatmap_z)

# Generate clustered heatmap
pheatmap(
  heatmap_z,
  scale = "none",
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "complete",
  annotation_col = sample_annotation,
  color = colorRampPalette(
    c("blue", "white", "red")
  )(100),
  show_rownames = TRUE,
  main = "Clustered Heatmap of Significant DEGs"
)
```

##### Pay attention to how the samples group together and compare this clustering pattern to what you observed in your earlier Spearman correlation analysis that used all genes.
- Comparing between Expressed Genes and Heatmap, we can observe that Spearman correlation analysis showing the similar pattern between genes and  the heat map is the comparision between all the cluster where we can see the difference color changed.




















