---
title: "Lab 4: Transcriptomics"
author: "Catherine"
format: html
editor: visual
---
<https://quarto.org>.
---

## 9.1.1 Question 1

##### Accessing the RNAseq data and sample data
```{r}
# Read the meta table 
meta_data <- read.table(
  system.file("extdata/param/targetsPE.txt", package = "systemPipeRdata"), 
  header = TRUE, 
  stringsAsFactors = FALSE
)

head(meta_data)

# Replace the filename
# Get the actual path to the FASTQ files directory
fq_path <- systemPipeRdata::pathList()$fastqdir

# Replace './data/' with the actual directory path in FileName1 and FileName2
meta_data$FileName1 <- sub("^\\./data/", paste0(fq_path, "/"), meta_data$FileName1)
meta_data$FileName2 <- sub("^\\./data/", paste0(fq_path, "/"), meta_data$FileName2)

# Verify the updated paths
head(meta_data[, c("FileName1", "FileName2")])


# Display the metadata table using DT package
library(DT)

# Render interactive table
datatable(
  meta_data, 
  options = list(scrollX = TRUE, pageLength = 5), 
  caption = "RNA-seq Sample Metadata Table"
)

```


## 9.2.1 Question 2

##### Read processing
```{r}
# Generate the QC report
library(Rqc)
library(ggplot2)

# Combine all forward and reverse file paths from your metadata table
all_files <- c(meta_data$FileName1, meta_data$FileName2)

# Run the overall Rqc quality assessment and generate the HTML report
qc_report <- rqc(path = dirname(all_files[1]), pattern = "fastq\\.gz$")
```

##### Generate 3 per-cycle Q-score box plot for files 1-12, 13-24, 25-36
```{r}
# Split the files into three equal groups of 12
files_1_12  <- all_files[1:12]
files_13_24 <- all_files[13:24]
files_25_36 <- all_files[25:36]

# Run rqcQA for each batch
qa_1 <- rqcQA(files_1_12)
qa_2 <- rqcQA(files_13_24)
qa_3 <- rqcQA(files_25_36)

# Generate per-cycle Q-score box plots with black and white theme
rqcCycleQualityBoxPlot(qa_1) + theme_bw()
rqcCycleQualityBoxPlot(qa_2) + theme_bw()
rqcCycleQualityBoxPlot(qa_3) + theme_bw()
```

##### Are there any samples with much longer reads than others?
- No, they share a uniform read length

##### What’s the shortest number of cycles?
- The shortest number of cycle is about 75 cycles

##### Generate 3 base call frequency plots for files 1-12, 13-24, 25-36
```{r}
# Generate base call frequency plots with black and white theme
rqcCycleBaseCallsPlot(qa_1) + theme_bw()
rqcCycleBaseCallsPlot(qa_2) + theme_bw()
rqcCycleBaseCallsPlot(qa_3) + theme_bw()
```

##### Does any sample have significant N calls?
- No, there is no signicicant N calls


## 9.2.2 Question 3
```{r}
dir.create("outputs/processed_fastq", 
           recursive = TRUE, 
           showWarnings = FALSE)

for (i in seq_along(meta_data$FileName1)) {
  
  fastqfiles <- c(meta_data$FileName1[i],
                  meta_data$FileName2[i])
  l <- length(stringr::str_split_1(meta_data$FileName1[i], "\\/"))
  
  
  outfiles <- c(paste0("outputs/processed_fastq/", stringr::str_split_1(meta_data$FileName1[i], "\\/")[l], "_processed.fastq.gz"),
                paste0("outputs/processed_fastq/", stringr::str_split_1(meta_data$FileName2[i], "\\/")[l], "_processed.fastq.gz"))
  
  # Remove existing output files
  file.remove(outfiles[file.exists(outfiles)])
  
  QuasR::preprocessReads(fastqfiles, outfiles,
                  nBases=1,
                  truncateEndBases=3,
                  Lpattern="GCCCGGGTAA",
                  minLength=40)
  
  meta_data$FileName1[i] <- outfiles[1]
  meta_data$FileName2[i] <- outfiles[2]
}
```

## 9.3.3 Question 4: Aligning the reads
```{r}
# Create output directories
dir.create(
  "outputs/sam_files",
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  "outputs/bam_files",
  recursive = TRUE,
  showWarnings = FALSE
)

# Check that all 18 paired-end processed FASTQ files exist
stopifnot(
  all(file.exists(meta_data$FileName1)),
  all(file.exists(meta_data$FileName2))
)

# Align each sample with HISAT2
for (i in seq_along(meta_data$FileName1)) {

  sample_name <- meta_data$SampleName[i]

  fastq_file1 <- meta_data$FileName1[i]
  fastq_file2 <- meta_data$FileName2[i]

  sam_output <- file.path(
    "outputs/sam_files",
    paste0(sample_name, ".sam")
  )

  bam_output <- file.path(
    "outputs/bam_files",
    paste0(sample_name, ".bam")
  )

  sorted_bam_output <- file.path(
    "outputs/bam_files",
    paste0(sample_name, "_sorted.bam")
  )

  log_output <- file.path(
    "outputs/bam_files",
    paste0(sample_name, "_hisat2.log")
  )

  message("Aligning ", sample_name, "...")

  # Run HISAT2
  hisat2_result <- system2(
    command = "hisat2",
    args = c(
      "-x", "outputs/hisat2_index/tair10_1_index",
      "-1", fastq_file1,
      "-2", fastq_file2,
      "-p", "8",
      "-S", sam_output
    ),
    stdout = TRUE,
    stderr = TRUE
  )

  # Save HISAT2 alignment statistics
  writeLines(
    hisat2_result,
    con = log_output
  )

  # Convert SAM to BAM
  system2(
    "samtools",
    args = c(
      "view",
      "-bS",
      sam_output,
      "-o",
      bam_output
    )
  )

  # Sort BAM file
  system2(
    "samtools",
    args = c(
      "sort",
      bam_output,
      "-o",
      sorted_bam_output
    )
  )

  # Index sorted BAM file
  system2(
    "samtools",
    args = c(
      "index",
      sorted_bam_output
    )
  )

  message("Finished ", sample_name)
}

# Confirm that all 18 SAM files were created
sam_files <- list.files(
  "outputs/sam_files",
  pattern = "\\.sam$"
)

length(sam_files)
```

## 9.3.4.1 Question 5: alignment stats
```{r}
library(DT)
library(ggplot2)
library(stringr)

# Directory containing HISAT2 log files
hisat2_logs_dir <- "outputs/bam_files"

# Find HISAT2 log files
log_files <- list.files(
  hisat2_logs_dir,
  pattern = "_hisat2\\.log$",
  full.names = TRUE
)

# Check that all 18 logs exist
if (length(log_files) != 18) {
  stop(
    paste(
      "Expected 18 HISAT2 log files, but found",
      length(log_files)
    )
  )
}

# Extract alignment percentages
align_df <- do.call(
  rbind,
  lapply(log_files, function(log_file) {

    log_text <- readLines(
      log_file,
      warn = FALSE
    )

    # Find the overall alignment rate
    alignment_line <- log_text[
      grepl(
        "overall alignment rate",
        log_text,
        ignore.case = TRUE
      )
    ]

    # Stop if the log does not contain a successful alignment
    if (length(alignment_line) == 0) {
      stop(
        paste(
          "No successful HISAT2 alignment rate found in:",
          basename(log_file)
        )
      )
    }

    # Extract percentage
    percent <- as.numeric(
      str_extract(
        alignment_line[1],
        "[0-9]+\\.?[0-9]*(?=%)"
      )
    )

    # Extract sample name
    sample_name <- sub(
      "_hisat2\\.log$",
      "",
      basename(log_file)
    )

    data.frame(
      samplename = sample_name,
      percent_aligned = percent,
      stringsAsFactors = FALSE
    )
  })
)

# Arrange samples in metadata order
align_df <- align_df[
  match(meta_data$SampleName, align_df$samplename),
]

# Reset row names
rownames(align_df) <- NULL

# Display results
align_df

# Interactive table
datatable(
  align_df,
  options = list(
    scrollX = TRUE,
    pageLength = 6
  ),
  caption = "HISAT2 Read Alignment Percentages per Sample"
)

# Boxplot of alignment percentages
ggplot(
  align_df,
  aes(
    x = "",
    y = percent_aligned
  )
) +
  geom_boxplot(
    fill = "steelblue",
    alpha = 0.7,
    outlier.color = "red",
    outlier.shape = 16
  ) +
  theme_bw() +
  labs(
    title = "Distribution of HISAT2 Read Alignment Percentages",
    x = "",
    y = "Percent Aligned (%)"
  )
```



