---
title: "Lab 4: Transcriptomics"
author: "Catherine"
format: html
editor: visual
---
<https://quarto.org>.
---

## 9.1.1 Question 1

##### Accessing the RNAseq data and sample data
```{r}
# Read the meta table 
meta_data <- read.table(
  system.file("extdata/param/targetsPE.txt", package = "systemPipeRdata"), 
  header = TRUE, 
  stringsAsFactors = FALSE
)

head(meta_data)

# Replace the filename
# Get the actual path to the FASTQ files directory
fq_path <- systemPipeRdata::pathList()$fastqdir

# Replace './data/' with the actual directory path in FileName1 and FileName2
meta_data$FileName1 <- sub("^\\./data/", paste0(fq_path, "/"), meta_data$FileName1)
meta_data$FileName2 <- sub("^\\./data/", paste0(fq_path, "/"), meta_data$FileName2)

# Verify the updated paths
head(meta_data[, c("FileName1", "FileName2")])


# Display the metadata table using DT package
library(DT)

# Render interactive table
datatable(
  meta_data, 
  options = list(scrollX = TRUE, pageLength = 5), 
  caption = "RNA-seq Sample Metadata Table"
)

```


## 9.2.1 Question 2

##### Read processing
```{r}
# Generate the QC report
library(Rqc)
library(ggplot2)

# Combine all forward and reverse file paths from your metadata table
all_files <- c(meta_data$FileName1, meta_data$FileName2)

# Run the overall Rqc quality assessment and generate the HTML report
qc_report <- rqc(path = dirname(all_files[1]), pattern = "fastq\\.gz$")
```

##### Generate 3 per-cycle Q-score box plot for files 1-12, 13-24, 25-36
```{r}
# Split the files into three equal groups of 12
files_1_12  <- all_files[1:12]
files_13_24 <- all_files[13:24]
files_25_36 <- all_files[25:36]

# Run rqcQA for each batch
qa_1 <- rqcQA(files_1_12)
qa_2 <- rqcQA(files_13_24)
qa_3 <- rqcQA(files_25_36)

# Generate per-cycle Q-score box plots with black and white theme
rqcCycleQualityBoxPlot(qa_1) + theme_bw()
rqcCycleQualityBoxPlot(qa_2) + theme_bw()
rqcCycleQualityBoxPlot(qa_3) + theme_bw()
```

##### Are there any samples with much longer reads than others?
- No, they share a uniform read length

##### What’s the shortest number of cycles?
- The shortest number of cycle is about 75 cycles

##### Generate 3 base call frequency plots for files 1-12, 13-24, 25-36
```{r}
# Generate base call frequency plots with black and white theme
rqcCycleBaseCallsPlot(qa_1) + theme_bw()
rqcCycleBaseCallsPlot(qa_2) + theme_bw()
rqcCycleBaseCallsPlot(qa_3) + theme_bw()
```

##### Does any sample have significant N calls?
- No, there is no signicicant N calls


## 9.2.2 Question 3
```{r}
dir.create("outputs/processed_fastq", recursive = TRUE)

for (i in seq_along(meta_data$FileName1)) {
  
  fastqfiles <- c(meta_data$FileName1[i], meta_data$FileName2[i])
  l <- length(stringr::str_split_1(meta_data$FileName1[i], "\\/"))
  
  outfiles <- c(paste0("outputs/processed_fastq/", stringr::str_split_1(meta_data$FileName1[i], "\\/")[l], "_processed.fastq.gz"),
                paste0("outputs/processed_fastq/", stringr::str_split_1(meta_data$FileName2[i], "\\/")[l], "_processed.fastq.gz"))
  
  QuasR::preprocessReads(fastqfiles, outfiles,
                  nBases=1,
                  truncateEndBases=3,
                  Lpattern="GCCCGGGTAA",
                  minLength=40)
  
  meta_data$FileName1[i] <- outfiles[1]
  meta_data$FileName2[i] <- outfiles[2]
}
```


## 9.3.3 Question 4: Aligning the reads













